Lessons learned from bugs in models of human history

### 2020-05-20 paper.tex
Edits to abstract and introduction, minor edits throughout

\aprcomment{title thoughts: \\
Reducing errors in demographic model design and implementation \\
Lessons learned from misspecified models of human history \\
Lessons learned from buggy models of human history \\
Lessons learned from bugs in population genetics models
}
\sgcomment{How to avoid simulating the wrong demographic model}

\abstract{
Simulation plays a central role in population genomics studies. Recent years
have seen rapid improvements in software efficiency that make it possible to simulate
large genomic regions for many individuals sampled from large numbers of populations.
\aprcomment{fix this sentence ->} The increase in complexity of possible demographic models also provides additional ways
that we can get their implementation wrong. Here we describe two errors made in
defining population genetic models using the \msprime\ coalescent simulator that have
found their way into the published record.
We discuss how these errors have affected analyses and suggest recommendations
for software developers and users to reduce the risk of such errors.
}

@@

\sgcomment{In the race to build more realistic and useful simulations of genetic diversity, 
scientific software developers often focus on computational speed and biological realism. 
User interface and application programming interfaces (API), however, also determine 
whether software is used, and whether it is used correctly.
 }

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used. Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which make it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is the Python API that is its primary interface, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces. In particular, programs like \ms\
required users to specify cryptic command line options to describe demographic
models \jkcomment{maybe include an example ms command line?}. Particularly for
the large models we use today, these are not intuitive or comprehensible for humans.
The Python API for \msprime\ is a great improvement, allowing the user to
state models in a documented and reproducible manner. \jkcomment{maybe show
the same model as described above in msprime notation?}. \sgcomment{Other software, such as SLiM, 
have developed advanced user interfaces and even dedicated programming languages 
to facilitate the specification of various modes of evolution.\cite{}
}

Implementing multi-population models of demographic history is still hard and error
prone, however. In this note we discuss two implementation errors that arose through
unfortunate design decisions in \msprime's demography API and which then found their way
into the scientific record. The first error has relatively mild effects on genetic diversity but was 
used in many publications, while the second error was used only once but had a large impact on the 
simulation results.  
In light of these implementation errors, we discuss improvements to \msprime's API motivated by these
discoveries and more general best practices for implementing and simulating complex multi-population
demography.

@@

To illustrate the demography API, \msprime\ included a description of a widely-used
three population Out-of-Africa model~\citep{gutenkunst2009inferring}
as part of its tutorial documentation. In this model (Fig.~\ref{fig:ooa_stats}A),
Eurasian (CEU and CHB) and African (YRI) populations split from each other in the deep past,
followed by a more recent split of European and Asian populations, with variable rates of
continuous migration between each of the populations. However, the implementation in the
\msprime\ tutorial was incorrect. Before the time of the split of African and Eurasian
populations, when there should have been just a single randomly mating population, migration was
allowed to occur between the ancestral population and a second population with size equal to
the Eurasian bottleneck size for all time into the past
(Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
the tutorial for \msprime\ version 0.3.0 and remained in the documentation for
around four years.

Fortunately, the effects of this error are subtle.
Population sizes and structure since the time of
the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
the correct and incorrect model. However, the ancient structure distorts the distribution
of early $T_{MRCAs}$. The extraneous ancient population increases the long-term effective 
population size, leading to heterozygosity in contemporary populations
increased by roughly $4\%$ over expectations from the correctly specified 
model (Fig.~\ref{fig:ooa_stats}E-F).

It is fortunate that the error in the model description has quite a
small effect because the tutorial code has been copied many times
and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications use simulations to test some inference
method which is not directly concerned with detailed demography,
and the model is simply used as an example of a complex population
history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
\cite{zhou2018popdemog} use the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
% TODO ask Franz about this
Finally, \cite{pfaffelhuber2020choose}
use simulations of the incorrect model demography to evalute
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
will have qualitative effect on their conclusions.

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
After these initial checks were made, however, the simulation
was performed without passing the parameter of demographic events,
so that the three populations
never merged as expected and remained separated with low levels of migration   (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: expected heterozygosity was more than
doubled in African populations but just $30-40\%$ the expected
levels in Eurasian populations when compared to diversity under the correctly specified model.

@@

The implementation of complex demographic models is error prone, and such errors
can have a large impact on downstream analysis and interpretation. 
The discovery and correction of the demographic models discussed here underscore
how API design choice can lead to the propagation of mistakes that are difficult
to notice. We therefore recommend the following steps to ensure more robust simulations.

First, the design of user interface and API for scientific software matters,
and many bugs can be prevented by using more intuitive interfaces.
Whereas the original \msprime\ required the user to pass
three separate parameters to specify a demographic model, \msprime~ver.~1.0
will introduce a Demography class, which wraps these three parameters.
Thus, instead of writing
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
\end{lstlisting}
with the error-inducing need to re-enter parameter information,
we would now write
\begin{lstlisting}[frame=single]
demography = msprime.Demography(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}

@@

We computed the expected allele frequency spectrum using
\texttt{moments}~ver.~1.0.3~\citep{jouganous2017inferring}, and LD-decay curves using
\texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other diversity statistics
were computed from the expected AFS and verified using branch statistics from the
tree sequence records of \msprime\ simulations~\citep{ralph2020efficiently}.
Demographic models were plotted using the \texttt{demography} package
(\url{https://github.com/apragsdale/demography}, ver.~0.0.3).
We used the original pipeline from~\citet{martin2017human} available from
\url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py}.
We updated the pipeline to run with the correct demographic parameters and more
recent versions of \msprime\ and \tskit\, and the updated pipeline is
available at \url{https://github.com/apragsdale/PRS}.
Data and python scripts to recreate Figures~1~and~2 can be found at
\url{https://github.com/jeromekelleher/msprime-model-errors}.
A full list of the GitHub repositories containing copies of the erroneous
model are also given here.

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/ooa_expected_stats.pdf}}
\caption{\textbf{Expected diversity statistics under the \citet{gutenkunst2009inferring} model}.
    \textbf{(A)} The correctly implemented model.
    \textbf{(B)} The incorrectly implemented model from the \msprime\ tutorial, with migration continuing
    into the past beyond the mass migration event with proportion $1$ from the ancestral population
    to the bottleneck population.
    \textbf{(C)} Marginal allele frequency spectra under the two models. Heterozygosity in the incorrect model
    is inflated by $~3.5\%$, though the general shape of the distributions are qualitatively similar.
    \textbf{(D)} Similarly, the increased heterozygosity leads to excess $D^2$, though the LD-decay is
    qualitatively similar between models.
    \textbf{(E-G)} True size history for each population plotted against the expected size history from
    the expected inverse coalescence rates.
}
\label{fig:ooa_stats}
\end{center}
\end{figure}

>>>>

%\aprcomment{title thoughts: \\
%Reducing errors in demographic model design and implementation \\
%Lessons learned from misspecified models of human history \\
%Lessons learned from buggy models of human history \\
%Lessons learned from bugs in population genetics models
%}

%\sgcomment{How to avoid simulating the wrong demographic model}

\abstract{
Simulation plays a central role in population genomics studies. Recent years
have seen rapid improvements in software efficiency that make it possible to simulate
large genomic regions for many individuals sampled from large numbers of populations.
As the complexity of demographic models we study grows, there is an ever-increasing
opportunity to introduce bugs in their implementation.
Here we describe two errors made in defining population genetic models using
the \msprime\ coalescent simulator that have found their way into the published record.
We discuss how these errors have affected downstream analyses and give recommendations
for software developers and users to reduce the risk of such errors.
}

@@

In the effort to build more realistic simulations of genetic diversity, 
scientific software developers often focus on computational speed and biological realism. 
Documentation and application programming interface (API), however, also determine 
whether software is used, and whether it is used correctly.

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is the Python API that is its primary interface, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ required users to specify cryptic command
line options to describe demographic models.
Particularly for the large models we use today, these are not
intuitive or comprehensible for humans.
The Python API for \msprime\ is a great improvement, allowing the user to
state models in a documented and reproducible manner.
\jkcomment{maybe show the OOA model as in both ms and msprime notation?}
%\sgcomment{Other software, such as SLiM, 
%have developed advanced user interfaces and even dedicated programming languages 
%to facilitate the specification of various modes of evolution.\cite{}
%}

Implementing multi-population models of demographic history is still hard and error
prone. In this note we discuss two implementation errors that arose through
unfortunate design decisions in \msprime's demography API and which then found their way
into the scientific record. The first error has relatively mild effects on genetic diversity but was 
used in many publications, while the second error was used only once but had a large impact on the 
simulation results.  
In light of these implementation errors, we discuss improvements to \msprime's API motivated by these
discoveries and more general best practices for implementing and simulating complex multi-population
demography.

@@

To illustrate the demography API, \msprime\ included a description of a widely-used
three population Out-of-Africa model~\citep{gutenkunst2009inferring}
as part of its tutorial documentation. In this model (Fig.~\ref{fig:ooa_stats}A),
Eurasian (CEU and CHB) and African (YRI) populations split from each other in the deep past,
followed by a more recent split of European and Asian populations, with variable rates of
continuous migration between each of the populations. However, the implementation in the
\msprime\ tutorial was incorrect. Before the time of the split of African and Eurasian
populations, when there should have been just a single randomly mating population, migration was
allowed to occur between the ancestral population and a second population with size equal to
the Eurasian bottleneck size for all time into the past
(Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
the tutorial for \msprime\~ver.~0.3.0 and remained in the documentation for
around four years.

Fortunately, the effects of this error are subtle.
Population sizes and structure since the time of
the split are unaffected, so differences in expected $F_{ST}$ are negligible between
the correct and incorrect model. However, the ancient structure distorts the distribution
of early TMRCAs. The extraneous ancient population increases the long-term effective 
population size, resulting in roughly $4\%$ excess heterozygosity in 
contemporary populations compared to the correctly specified model
(Fig.~\ref{fig:ooa_stats}C-F).

It is fortunate that the error in the model description has quite a
small effect because the tutorial code has been copied many times
and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications used simulations to test some inference
method and were not directly concerned with the details of the demography,
and the model was simply used as an example of a complex population
history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
\cite{zhou2018popdemog} used the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
% TODO ask Franz about this
Finally, \cite{pfaffelhuber2020choose}
used simulations of the incorrect model demography to evaluate
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
had any qualitative effect on their conclusions.

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
After these initial checks were made, however, the simulation
was performed without passing the parameter of demographic events,
so that the three populations
never merged as expected and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: heterozygosity was more than
doubled in African populations but just $30-40\%$ the expected
levels in Eurasian populations when compared to diversity under the correctly specified model.

@@

The implementation of complex demographic models is error prone, and such errors
can have a large impact on downstream analyses and interpretation. 
The discovery and correction of the demographic models discussed here underscore
how API design choice can lead to the propagation of mistakes that are difficult
to notice. We therefore recommend the following steps to ensure more robust simulations.

First, the design of user interface and API for scientific software matters,
and many bugs can be prevented by using more intuitive interfaces.
Whereas the original \msprime\ API required the user to pass
three separate parameters to specify a demographic model, \msprime~ver.~1.0
will introduce a Demography class, which wraps these three parameters.
Thus, instead of writing
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
\end{lstlisting}
with the error-inducing need to re-enter parameter information,
we would now write
\begin{lstlisting}[frame=single]
demography = msprime.Demography(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}

@@

We computed the expected allele frequency spectrum using
\texttt{moments}~ver.~1.0.3~\citep{jouganous2017inferring}, and LD-decay curves using
\texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other diversity statistics
were computed from the expected AFS and verified using branch statistics from the
tree sequence records of \msprime\ simulations using \tskit~ver.~0.2.3~\citep{ralph2020efficiently}.
Demographic models were plotted using the \texttt{demography} package
(\url{https://github.com/apragsdale/demography}, ver.~0.0.3).
We used the original pipeline from~\citet{martin2017human} available from
\url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py}.
We updated the pipeline to run with the correct demographic parameters and more
recent versions of \msprime\ and \tskit, and the updated pipeline is
available at \url{https://github.com/apragsdale/PRS}.
Data and python scripts to recreate Figures~1~and~2 can be found at
\url{https://github.com/jeromekelleher/msprime-model-errors}.
A full list of the GitHub repositories containing copies of the erroneous
model are also given here.

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/ooa_expected_stats.pdf}}
\caption{\textbf{Expected diversity statistics under the \citet{gutenkunst2009inferring} model}.
    \textbf{(A)} The correctly implemented model. Dashed arrows depict continuous migration.
    \textbf{(B)} The incorrectly implemented model from the \msprime\ tutorial, with migration continuing
    into the past beyond the mass migration event with proportion $1$ from the ancestral population
    to the bottleneck population.
    \textbf{(C)} Marginal allele frequency spectra under the two models. Heterozygosity in the incorrect model
    is inflated by $~3.5\%$, though the general shape of the distributions are qualitatively similar.
    \textbf{(D)} Similarly, the increased heterozygosity leads to excess $D^2$, though the LD-decay is
    qualitatively similar between models.
    \textbf{(E-G)} True size history for each population plotted against the expected size history from
    the expected inverse coalescence rates.
}
\label{fig:ooa_stats}
\end{center}
\end{figure}
